---
title: "TNF Screen Analysis"
author: "pdevant"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 7,
  fig.height = 6,
  fig.align = "center"
)

library(ggplot2)
library(ggrepel)
library(dplyr)
library(grid)
library(patchwork)
```

## Read and Merge Gene Summary Files

This section reads the donor 1 and donor 2 MAGeCK gene summary outputs and merges them by gene `id`.

```{r read-and-merge}
df_D1 <- read.csv("/Users/pdevant/results_TNF2/D1_hi_vs_low_2025-05-23.gene_summary.csv", header = TRUE)

df_D2 <- read.csv("/Users/pdevant/results_TNF2/D2_hi_vs_low_2025-05-23.gene_summary.csv", header = TRUE)

merged_df <- merge(
  df_D1,
  df_D2,
  by = "id",
  suffixes = c("_D1", "_D2")
)
```

## Read Paired Analysis

This section loads the paired analysis used for the paired FDR values and the paired volcano plot.

```{r read-paired}
df_D12_TNF <- read.csv("/Users/pdevant/results_TNF_paired_perm100000/Paired_hi_vs_low_2026-02-15.gene_summary.csv", header = TRUE)

## Calculate z-scores using scale function

# Run scale() on neg.lfc and add as new column
df_D12_TNF$z_score <- as.numeric(scale(df_D12_TNF$`neg.lfc`))

## Add FDR column
df_D12_TNF$FDR <- ifelse(df_D12_TNF$pos.lfc > 0, df_D12_TNF$pos.fdr, df_D12_TNF$neg.fdr)

## Add extra columns that stored -log10 of FDRs (positive and negative)
df_D12_TNF$neglog10_FDR <- -log10(df_D12_TNF$FDR)
```

## Replot donor-donor correlation plot, but use z-scores and color by paired FDR

Here, donor 1 and donor 2 are z-scored independently, then paired FDR values from the paired analysis are added to the merged donor table.

```{r donor-correlation-z}
merged_df$z_score_D1 <- as.numeric(scale(merged_df$pos.lfc_D1))
merged_df$z_score_D2 <- as.numeric(scale(merged_df$pos.lfc_D2))

merged_df <- merged_df %>%
  left_join(
    df_D12_TNF %>% select(id, FDR),
    by = "id"
  ) %>%
  rename(FDR_paired = FDR)

merged_df$group <- with(merged_df, ifelse(
  z_score_D1 > 3 & z_score_D2 > 3, "GroupA",
  ifelse(z_score_D1 < -3 & z_score_D2 < -3, "GroupB", "Other")
))

label_df <- subset(merged_df, group %in% c("GroupA", "GroupB"))

x_limits <- range(merged_df$z_score_D1, na.rm = TRUE)
y_limits <- range(merged_df$z_score_D2, na.rm = TRUE)

x_pad <- diff(x_limits) * 0.05
y_pad <- diff(y_limits) * 0.05

if (!is.finite(x_pad) || x_pad == 0) x_pad <- 0.5
if (!is.finite(y_pad) || y_pad == 0) y_pad <- 0.5
```

```{r donor-correlation-z-plot}
set.seed(1)

p3_z_fig <- ggplot(merged_df,
                   aes(x = z_score_D1,
                       y = z_score_D2,
                       color = FDR_paired)) +
  geom_point(size = 2, alpha = 0.55) +
  scale_color_viridis_c(
    option = "D",
    direction = 1,
    name = "Paired FDR",
    limits = c(0, 1),
    oob = scales::squish,
    guide = guide_colorbar(
      barheight = unit(35, "mm"),
      barwidth  = unit(5, "mm"),
      ticks.colour = "black",
      frame.colour = "black"
    )
  ) +
  scale_x_continuous(limits = c(x_limits[1] - x_pad, x_limits[2] + x_pad)) +
  scale_y_continuous(limits = c(y_limits[1] - y_pad, y_limits[2] + y_pad)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_text_repel(
    data = label_df,
    aes(label = id),
    seed = 1,
    size = 3,
    fontface = "italic",
    box.padding   = 0.6,
    point.padding = 0.7,
    force         = 10,
    force_pull    = 2,
    max.overlaps  = Inf,
    max.iter      = 50000,
    max.time      = 5,
    min.segment.length = 0,
    segment.color = "gray35",
    segment.size  = 0.35,
    segment.alpha = 0.9
  ) +
  coord_fixed(ratio = 1, clip = "off") +
  labs(
    title = "Donor-Donor correlation (z-scores)",
    x = "z-score (Donor 1)",
    y = "z-score (Donor 2)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 11),
    legend.position = "right",
    plot.margin = margin(10, 25, 10, 10)
  )

print(p3_z_fig)
```

## Plotting of Volcano plot D1 and D2 paired analysis

This section reproduces the paired volcano plot using z-scores, an FDR cutoff of 0.25, and separate color bins for z-scores between 1 and 3 versus greater than 3. The split x-axis is used to show the far-left outlier region together with the main distribution.

```{r volcano-z-bins}
# Thresholds
FDR_threshold <- 0.25
z_lo <- 1
z_hi <- 3

# Define 4 color bins (only if FDR passes)
df_D12_TNF$ColorGroup <- with(df_D12_TNF, ifelse(
  FDR < FDR_threshold & z_score >= z_hi, "Up_hi",
  ifelse(FDR < FDR_threshold & z_score >= z_lo & z_score < z_hi, "Up_lo",
         ifelse(FDR < FDR_threshold & z_score <= -z_hi, "Down_hi",
                ifelse(FDR < FDR_threshold & z_score <= -z_lo & z_score > -z_hi, "Down_lo",
                       "Neutral")
         )
  )
))

# Label genes that pass FDR AND |z| >= 1
label_df <- df_D12_TNF %>%
  filter(FDR < FDR_threshold, abs(z_score) >= z_lo)

# Volcano with a true x-axis break (two panels) + consistent x-scale + FDR label
x_left <- c(-16, -14.5)
x_main <- c(-8, 8)

y_fdr <- -log10(FDR_threshold)

df_left <- df_D12_TNF %>%
  filter(z_score >= x_left[1], z_score <= x_left[2])

df_main <- df_D12_TNF %>%
  filter(z_score >= x_main[1], z_score <= x_main[2])

label_left <- label_df %>%
  filter(z_score >= x_left[1], z_score <= x_left[2])

if ("TNF" %in% label_left$id) {
  label_left <- label_left %>% filter(id == "TNF")
} else if (nrow(label_left) > 0) {
  label_left <- label_left %>% slice_max(order_by = abs(z_score), n = 1)
}

label_main <- label_df %>%
  filter(z_score >= x_main[1], z_score <= x_main[2])

y_top <- max(df_D12_TNF$neglog10_FDR, na.rm = TRUE) + 0.7

col_scale <- scale_color_manual(values = c(
  "Down_hi" = "#08519C",
  "Down_lo" = "#9ECAE1",
  "Neutral" = "gray80",
  "Up_lo"   = "#FCBBA1",
  "Up_hi"   = "#CB181D"
))

base_theme <- theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    plot.margin = margin(10, 10, 10, 10)
  )

left_breaks <- c(-15)

main_breaks <- c(pretty(x_main), -8, 8, -z_hi, -z_lo, 0, z_lo, z_hi)

x_fdr_label <- x_main[1] + 0.2
```

```{r volcano-z-bins-plot}
p_left <- ggplot(df_left, aes(x = z_score, y = neglog10_FDR)) +
  geom_point(aes(color = ColorGroup), alpha = 0.8, size = 2) +
  col_scale +
  geom_hline(yintercept = y_fdr, linetype = "dashed", color = "grey60") +
  geom_text_repel(
    data = label_left,
    aes(label = id),
    size = 3,
    box.padding = 0.4,
    point.padding = 0.2,
    force = 3,
    max.overlaps = Inf,
    segment.color = "grey60",
    segment.size = 0.3
  ) +
  coord_cartesian(ylim = c(0, y_top), clip = "off") +
  scale_x_continuous(breaks = left_breaks, limits = x_left) +
  labs(x = NULL, y = "-Log10(FDR)") +
  base_theme

p_main <- ggplot(df_main, aes(x = z_score, y = neglog10_FDR)) +
  geom_point(aes(color = ColorGroup), alpha = 0.8, size = 2) +
  col_scale +
  geom_vline(
    xintercept = c(-z_hi, -z_lo, z_lo, z_hi),
    linetype = "dashed",
    color = "grey60"
  ) +
  geom_hline(yintercept = y_fdr, linetype = "dashed", color = "grey60") +
  annotate(
    "text",
    x = x_fdr_label,
    y = y_fdr,
    label = paste0("FDR = ", FDR_threshold),
    hjust = 0,
    vjust = -0.6,
    size = 3,
    color = "grey40"
  ) +
  geom_text_repel(
    data = label_main,
    aes(label = id),
    size = 3,
    box.padding = 0.5,
    point.padding = 0.3,
    force = 3,
    max.overlaps = 50,
    segment.color = "grey60",
    segment.size = 0.3
  ) +
  coord_cartesian(ylim = c(0, y_top), clip = "off") +
  scale_x_continuous(breaks = main_breaks, limits = x_main) +
  labs(x = "LFC z-score", y = NULL) +
  base_theme +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y  = element_blank()
  )

left_span <- diff(x_left)
main_span <- diff(x_main)

zVolcanoTNF <- p_left + p_main +
  plot_layout(widths = c(left_span, main_span))

print(zVolcanoTNF)
```
